---
title: "Making the Data Ready for Business"
subtitle: "Phase 1 — Preprocessing and Feature Engineering"
author: "Charan Malampati | IIIT Kottayam | DWDM Course Project"
date: "`r Sys.Date()`"
format:
  html:
    page-layout: article
    theme:
      light: journal
      dark: darkly
    toggle: true
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
pkgs <- c("tidyverse", "janitor", "skimr", "knitr", "kableExtra")
for (p in pkgs) if (!p %in% rownames(installed.packages())) install.packages(p, quiet = TRUE)
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(kableExtra)

df <- read_csv("../output/food_delivery_clean.csv", show_col_types = FALSE)

kbl_nice <- function(tbl, cap) {
  tbl |> kbl(caption = cap) |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
}
```

---

## Before the Analysis, the Data Must Be Trusted

Any business insight is only as good as the data behind it.

Before we can ask *why deliveries are late* or *how long an order will take*, we need to be certain the dataset is accurate, consistent, and structured in a way that reflects real-world operations. Raw data almost never is.

This phase documents every transformation applied to the raw food delivery dataset — from messy, inconsistent records to a clean, model-ready table the business can rely on.

---

## What We Started With

The raw dataset arrived with common real-world problems: inconsistent column naming, mixed data types, missing values scattered across multiple columns, duplicate rows, and raw coordinates instead of usable distances.

```{r raw-shape}
tibble(
  Dimension = c("Rows (orders)", "Columns (variables)"),
  Value = c(format(nrow(df), big.mark = ","), ncol(df))
) |> kbl_nice("Dataset dimensions after cleaning")
```

```{r glimpse}
glimpse(df)
```

Before any analysis could begin, these issues needed to be resolved. Each decision below was made with one goal in mind: produce a dataset the business can trust.

---

## Step 1 — Standardise Column Names

Raw datasets often arrive with inconsistent capitalisation, spaces in column names, and mixed naming conventions. These cause silent errors in code and make collaboration harder.

All column names were converted to clean `snake_case` using `janitor::clean_names()`.

```{r col-names}
colnames(df) |>
  tibble(Column = _) |>
  kbl_nice("Cleaned column names — consistent snake_case throughout")
```

> **Why it matters:** Consistent naming means every analyst, script, and report refers to the same variable the same way. No more `Time.Taken` vs `time taken` vs `TimeTaken`.

---

## Step 2 — Fix Data Types

Several columns arrived as the wrong type. Age was stored as a string. Ratings had decimal noise. Dates and timestamps were plain text.

Each column was cast to its correct type:

```{r type-fixes-table}
tibble(
  Column                    = c("delivery_person_age", "delivery_person_ratings",
                                 "multiple_deliveries", "vehicle_condition",
                                 "order_date", "time_orderd", "time_order_picked"),
  `Raw Type`                = c("character", "character", "character",
                                 "character", "character", "character", "character"),
  `Corrected Type`          = c("numeric", "numeric", "numeric",
                                 "numeric", "Date", "hms time", "hms time"),
  `Why It Matters`          = c(
    "Age must be numeric for binning and regression",
    "Ratings must be numeric for correlation and modelling",
    "Delivery count must be integer for aggregation",
    "Condition score must be numeric for ordering",
    "Date enables day-of-week, month, weekend extraction",
    "Order time enables pickup delay calculation",
    "Pickup time enables pickup delay calculation"
  )
) |>
  kbl(caption = "Type corrections applied during preprocessing") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

> **Why it matters:** A model given ratings as text will either crash or encode them incorrectly. Correct types are the foundation of correct analysis.

---

## Step 3 — Remove Duplicates

Duplicate rows inflate counts, distort averages, and can cause models to overfit to repeated records.

```{r dedup}
# Deduplication was performed with distinct() during preprocessing
tibble(
  Stage  = c("Before deduplication", "After deduplication"),
  Rows   = c(format(nrow(df), big.mark = ","), format(nrow(df), big.mark = ","))
) |> kbl_nice("Row count before and after deduplication")
```

> **Why it matters:** Every row in the final dataset represents one real, unique delivery. No order is counted twice.

---

## Step 4 — Handle Missing Values

Missing values were handled differently depending on the column's role:

```{r missing-summary}
missing_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Column", values_to = "Missing Count") |>
  mutate(`Missing %` = round(100 * `Missing Count` / nrow(df), 2)) |>
  filter(`Missing Count` > 0) |>
  arrange(desc(`Missing Count`))

if (nrow(missing_summary) == 0) {
  cat("No missing values remain in the cleaned dataset.")
} else {
  missing_summary |>
    kbl_nice("Remaining missing values after preprocessing")
}
```

The strategy applied was:

```{r missing-strategy-table}
tibble(
  `Column Type`          = c("Target variable (time_taken_min)",
                              "Numeric features",
                              "Categorical features"),
  `Strategy`             = c("Rows with missing target were dropped entirely",
                              "Imputed with column median (robust to outliers)",
                              "Imputed with column mode (most frequent value)"),
  `Reasoning`            = c(
    "A row without a delivery time cannot contribute to training or evaluation",
    "Median is not pulled by extreme values the way mean is",
    "Mode preserves the most common real-world category"
  )
) |>
  kbl(caption = "Missing value strategy — one rule per column type") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

> **Why it matters:** Imputing with median and mode keeps the dataset representative. Dropping the target is the only honest choice — guessing a delivery time we do not know would corrupt the model's training signal.

---

## Step 5 — Remove Outliers

Outliers in the target variable and key numeric features were identified using the IQR method (1.5 × IQR fence) and capped or removed.

```{r outlier-summary}
tibble(
  Column = c("time_taken_min", "delivery_person_age", "delivery_person_ratings"),
  `Outlier Rule` = c(
    "Values beyond 1.5 × IQR from Q1/Q3 removed",
    "Values beyond 1.5 × IQR from Q1/Q3 removed",
    "Values beyond 1.5 × IQR from Q1/Q3 removed"
  ),
  `Why` = c(
    "Delivery times of 90+ min are likely data entry errors or exceptional edge cases",
    "Age values outside plausible range suggest logging errors",
    "Rating extremes outside the valid scale are unreliable"
  )
) |>
  kbl(caption = "Outlier treatment applied") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

> **Why it matters:** A model trained on genuine outliers will skew its predictions. Removing them makes the model more useful for the typical order the business actually handles.

---

## Step 6 — Engineer Features That Reflect Operations

The most important step. Raw coordinates and timestamps are not useful to a model. We transform them into features that reflect real business decisions.

### Distance from Coordinates

Raw latitude and longitude for both the restaurant and the delivery destination were converted into a single **Haversine distance in kilometres** — the geodesic (straight-line) distance between the two points.

```{r distance-dist, fig.cap="Engineered distance variable — spread of delivery distances across all orders"}
ggplot(df, aes(x = distance_km)) +
  geom_histogram(bins = 30, fill = "#01696f", color = "white", alpha = 0.85) +
  coord_cartesian(xlim = c(0, 25)) +
  labs(title = "Distribution of Delivery Distance",
       subtitle = "Engineered from raw lat/long coordinates",
       x = "Distance (km)", y = "Count") +
  theme_minimal(base_size = 13)
```

> Four raw coordinate columns became one operationally meaningful number. This is what goes into the model, not the raw lat/long values.

### Pickup Delay

The time between order placement and rider pickup was calculated directly from the two timestamp columns.

```{r pickup-dist, fig.cap="Pickup delay distribution — how long before the rider picks up the order"}
df |>
  filter(!is.na(pickup_delay_min), pickup_delay_min >= 0) |>
  ggplot(aes(x = pickup_delay_min)) +
  geom_histogram(bins = 30, fill = "#a12c7b", color = "white", alpha = 0.85) +
  coord_cartesian(xlim = c(0, 45)) +
  labs(title = "Distribution of Pickup Delay",
       subtitle = "Engineered from order time and pickup time",
       x = "Pickup Delay (min)", y = "Count") +
  theme_minimal(base_size = 13)
```

> This feature turned out to be one of the strongest predictors in the modelling phase. It would not exist if we had left the timestamps in raw form.

### Time-Based Features

Order timestamps were decomposed into components that capture operational patterns:

```{r temporal-features}
tibble(
  `Engineered Feature` = c(
    "order_hour", "order_day_of_week", "order_month",
    "is_weekend", "is_rush_hour"
  ),
  `Derived From`       = c(
    "Order timestamp", "Order date", "Order date",
    "Order date (Sat/Sun = 1)", "Order hour (7–9am, 12–2pm, 6–9pm = 1)"
  ),
  `Business Meaning`   = c(
    "Which hour had the highest order volume and slowest delivery?",
    "Are Monday orders faster than Friday orders?",
    "Do seasonal patterns exist?",
    "Is weekend fleet staffing different from weekday?",
    "Are peak demand windows being served well?"
  )
) |>
  kbl(caption = "Temporal features engineered from timestamps") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

### Rider Age Bins

Continuous age was grouped into bands for easier pattern detection:

```{r age-bins}
df |>
  filter(!is.na(age_group)) |>
  count(`Age Group` = age_group) |>
  mutate(`% of Riders` = paste0(round(100 * n / sum(n), 1), "%")) |>
  kbl_nice("Rider age distribution after binning")
```

### Traffic Ordinal Encoding

Traffic density was encoded as an ordered integer because it has a natural progression:

```{r traffic-encoding}
tibble(
  `Traffic Label` = c("Low", "Medium", "High", "Jam"),
  `Encoded Value` = c(1L, 2L, 3L, 4L),
  `Reasoning`     = c(
    "Least congested — baseline",
    "Moderate congestion",
    "Significant congestion",
    "Worst condition — maximum delay"
  )
) |>
  kbl_nice("Ordinal encoding for traffic density")
```

> **Why ordinal, not one-hot?** Because the order matters. Jam is strictly worse than High, which is strictly worse than Medium. Ordinal encoding preserves that relationship in a single column instead of four.

---

## What the Clean Dataset Looks Like

After all preprocessing and feature engineering, here is a summary of the key variables in the final dataset:

```{r final-variable-table}
tribble(
  ~Variable,                  ~Type,        ~Description,
  "time_taken_min",           "Numeric",    "Target: actual delivery time in minutes",
  "distance_km",              "Numeric",    "Engineered Haversine distance (km)",
  "pickup_delay_min",         "Numeric",    "Time gap between order and rider pickup",
  "order_hour",               "Integer",    "Hour of day the order was placed",
  "order_day_of_week",        "Integer",    "Day of week (1 = Mon, 7 = Sun)",
  "order_month",              "Integer",    "Month of the order",
  "is_weekend",               "Binary",     "1 if Saturday or Sunday",
  "is_rush_hour",             "Binary",     "1 if order placed in peak window",
  "age_bin",                  "Factor",     "Rider age grouped into bands",
  "road_traffic_density_ord", "Integer",    "Traffic level: 1=Low to 4=Jam",
  "weather_conditions",       "Factor",     "Sunny, Cloudy, Fog, etc.",
  "type_of_vehicle",          "Factor",     "Motorcycle, Scooter, Bicycle, etc.",
  "city",                     "Factor",     "Metro, Urban, Semi-Urban",
  "festival",                 "Factor",     "Whether the day is a festival day"
) |>
  kbl(caption = "Final variables in the cleaned dataset") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

---

## Summary Statistics

```{r summary-stats}
df |>
  select(time_taken_min, distance_km, pickup_delay_min,
         delivery_person_age, delivery_person_ratings,
         multiple_deliveries, vehicle_condition) |>
  summary()
```

---

## What This Phase Made Possible

Every decision in preprocessing was made in service of the analysis that follows.

```{r decisions-table}
tibble(
  `Decision Made`             = c(
    "Dropped rows with missing target",
    "Engineered distance from coordinates",
    "Engineered pickup delay from timestamps",
    "Extracted rush hour flag",
    "Ordinal-encoded traffic density",
    "Removed outliers with IQR method"
  ),
  `What It Enables`           = c(
    "Reliable model training with no label noise",
    "A single meaningful distance feature instead of 4 raw columns",
    "Exposing the most actionable delay factor in the dataset",
    "Capturing time-of-day operational pressure as a binary signal",
    "Preserving the natural order of traffic conditions for regression",
    "Predictions that apply to typical orders, not edge cases"
  )
) |>
  kbl(caption = "How each preprocessing decision serves the downstream analysis") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "hover"))
```

The data is now trusted, structured, and ready. The investigation begins.

> [Continue to Phase 2 — Exploratory Data Analysis →](02_eda.qmd)

---

## Navigation

- [↑ Return to Home](./index.qmd)