---
title: "Phase 1 - Preprocessing and Feature Engineering"
format:
  html:
    page-layout: article
    theme:
      light: journal
      dark: darkly
    toggle: true
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
---

## Objective

This phase converts the raw food delivery dataset into a clean, model-ready dataset by fixing data types, handling missing values, removing duplicates and outliers, and creating engineered features.

## Libraries

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(kableExtra)
```

## Load cleaned dataset

```{r}
df <- read_csv("../output/food_delivery_clean.csv", show_col_types = FALSE)
dim(df)
```

## Dataset overview

```{r}
glimpse(df)
```

## Final variables

```{r}
tribble(
  ~Variable, ~Purpose,
  "time_taken_min", "Target variable: delivery time in minutes",
  "distance_km", "Engineered Haversine distance between restaurant and destination",
  "pickup_delay_min", "Time gap between order placement and pickup",
  "order_hour", "Hour of order",
  "order_day_of_week", "Day of week extracted from order date",
  "order_month", "Month extracted from order date",
  "is_weekend", "Weekend indicator",
  "age_group", "Binned delivery partner age",
  "is_rush_hour", "Peak-time indicator",
  "road_traffic_density_ord", "Ordinal traffic encoding"
) |>
  kbl(caption = "Important processed and engineered variables") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Missing values after preprocessing

```{r}
missing_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") |>
  mutate(missing_pct = round(100 * missing_count / nrow(df), 2)) |>
  arrange(desc(missing_count))

missing_summary |>
  filter(missing_count > 0) |>
  kbl(caption = "Remaining missing values after preprocessing") |>
  kable_styling(full_width = FALSE)
```

## Summary statistics

```{r}
df |>
  select(time_taken_min, distance_km, pickup_delay_min,
         delivery_person_age, delivery_person_ratings,
         multiple_deliveries, vehicle_condition) |>
  summary()
```

## Distribution of engineered features

### Distance

```{r}
ggplot(df, aes(x = distance_km)) +
  geom_histogram(bins = 30, fill = "#01696f", color = "white") +
  labs(title = "Distribution of Delivery Distance", x = "Distance (km)", y = "Count") +
  theme_minimal(base_size = 13)
```

### Pickup delay

```{r}
ggplot(df, aes(x = pickup_delay_min)) +
  geom_histogram(bins = 30, fill = "#a12c7b", color = "white") +
  labs(title = "Distribution of Pickup Delay", x = "Pickup Delay (min)", y = "Count") +
  theme_minimal(base_size = 13)
```

## Key preprocessing notes

- Raw IDs were retained in the cleaned dataset but excluded later from modelling.
- Traffic was ordinally encoded because it has natural order: Low < Medium < High < Jam.
- Time-based features were created from date and order/pickup timestamps.
- Distance was engineered from latitude and longitude because it is more meaningful than raw coordinates alone.

## Navigation

Continue to [EDA](02_eda.qmd)

