---
title: "Phase 2 - Exploratory Data Analysis"
format:
  html:
    page-layout: article
    theme:
      light: journal
      dark: darkly
    toggle: true
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
---

## Objective

This phase explores how delivery time changes with distance, traffic, weather, city type, vehicle type, and pickup delay.

## Libraries

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Load cleaned dataset

```{r}
df <- read_csv("../output/food_delivery_clean.csv", show_col_types = FALSE)
dim(df)
```

## Dataset snapshot

```{r}
glimpse(df)
```

## Summary statistics

```{r}
df |>
  summarise(
    mean_time = mean(time_taken_min, na.rm = TRUE),
    median_time = median(time_taken_min, na.rm = TRUE),
    min_time = min(time_taken_min, na.rm = TRUE),
    max_time = max(time_taken_min, na.rm = TRUE),
    sd_time = sd(time_taken_min, na.rm = TRUE),
    mean_distance = mean(distance_km, na.rm = TRUE),
    mean_pickup_delay = mean(pickup_delay_min, na.rm = TRUE)
  ) |>
  kbl(caption = "Summary of key numeric variables") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Missing values

```{r}
missing_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") |>
  mutate(missing_pct = round(100 * missing_count / nrow(df), 2)) |>
  arrange(desc(missing_count))

missing_summary |>
  filter(missing_count > 0) |>
  kbl(caption = "Missing values after preprocessing") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Distribution of delivery time

```{r}
ggplot(df, aes(x = time_taken_min)) +
  geom_histogram(bins = 30, fill = "#01696f", color = "white") +
  geom_density(aes(y = after_stat(count)), color = "#a12c7b", linewidth = 1) +
  labs(
    title = "Distribution of Delivery Time",
    x = "Time Taken (min)",
    y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

## Correlation heatmap

```{r}
num_df <- df |>
  select(where(is.numeric)) |>
  select(-any_of("id"))

cor_mat <- cor(num_df, use = "pairwise.complete.obs")

cor_long <- as.data.frame(cor_mat) |>
  tibble::rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(cor_long, aes(var1, var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#01696f", midpoint = 0) +
  coord_fixed() +
  labs(
    title = "Correlation Heatmap of Numeric Variables",
    x = NULL,
    y = NULL,
    fill = "Corr"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Distance versus time

```{r}
ggplot(df, aes(x = distance_km, y = time_taken_min)) +
  geom_point(alpha = 0.35, color = "#01696f") +
  geom_smooth(method = "lm", se = TRUE, color = "#a12c7b") +
  labs(
    title = "Delivery Time vs Distance",
    x = "Distance (km)",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13)
```

## Traffic effect

```{r}
ggplot(df, aes(x = road_traffic_density, y = time_taken_min, fill = road_traffic_density)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  labs(
    title = "Delivery Time by Traffic Density",
    x = "Traffic Density",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Weather effect

```{r}
ggplot(df, aes(x = weather_conditions, y = time_taken_min, fill = weather_conditions)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  labs(
    title = "Delivery Time by Weather",
    x = "Weather",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

## City and festival effect

```{r}
ggplot(df, aes(x = city, y = time_taken_min, fill = city)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  labs(
    title = "Delivery Time by City",
    x = "City",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r}
ggplot(df, aes(x = festival, y = time_taken_min, fill = festival)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  labs(
    title = "Delivery Time by Festival",
    x = "Festival",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Pickup delay effect

```{r}
ggplot(df, aes(x = pickup_delay_min, y = time_taken_min)) +
  geom_point(alpha = 0.35, color = "#01696f") +
  geom_smooth(method = "lm", se = TRUE, color = "#a12c7b") +
  labs(
    title = "Pickup Delay vs Delivery Time",
    x = "Pickup Delay (min)",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13)
```

## Age and rating distributions

```{r}
ggplot(df, aes(x = delivery_person_age)) +
  geom_histogram(bins = 25, fill = "#01696f", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Count") +
  theme_minimal(base_size = 13)
```

```{r}
ggplot(df, aes(x = delivery_person_ratings)) +
  geom_histogram(bins = 25, fill = "#a12c7b", color = "white") +
  labs(title = "Distribution of Ratings", x = "Ratings", y = "Count") +
  theme_minimal(base_size = 13)
```

## Vehicle type effect

```{r}
ggplot(df, aes(x = type_of_vehicle, y = time_taken_min, fill = type_of_vehicle)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  labs(
    title = "Delivery Time by Vehicle Type",
    x = "Vehicle Type",
    y = "Time Taken (min)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Category summaries

```{r}
weather_summary <- df |>
  group_by(weather_conditions) |>
  summarise(
    count = n(),
    avg_time = mean(time_taken_min, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_time))

weather_summary |>
  kbl(caption = "Average delivery time by weather condition") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r}
city_summary <- df |>
  group_by(city) |>
  summarise(
    count = n(),
    avg_time = mean(time_taken_min, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_time))

city_summary |>
  kbl(caption = "Average delivery time by city") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Key observations

- Delivery time generally increases with distance.
- Heavy traffic and jam conditions increase delivery time more than low traffic.
- Pickup delay also contributes directly to final delivery time.
- Weather and city category create visible differences in delivery performance.

## Navigation

- Back to [Preprocessing](01_preprocessing.qmd)
- Continue to [Modelling](03_modelling.qmd)