---
title: "Phase 2 - EDA: Why Are Deliveries Getting Late?"
subtitle: "An Exploratory Investigation into Food Delivery Operations"
format:
  html:
    page-layout: article
    theme:
      light: journal
      dark: darkly
    toggle: true
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
    warning: false
    message: false
    echo: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
pkgs <- c("tidyverse", "knitr", "kableExtra", "scales", "ggplot2")
for (p in pkgs) if (!p %in% rownames(installed.packages())) install.packages(p, quiet = TRUE)
library(tidyverse); library(knitr); library(kableExtra); library(scales); library(ggplot2)

df <- read_csv("../output/food_delivery_clean.csv", show_col_types = FALSE)

kbl_nice <- function(tbl, cap) {
  tbl |> kbl(caption = cap) |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
}
```

---

## The Brief

A food delivery company has handed you their operational data and one question:

> *"Our delivery times are inconsistent. Customers are complaining. Figure out what is going on."*

You have **`r format(nrow(df), big.mark = ",")` delivery records**. Each row is one order — from placement to doorstep. Your job is to explore the data, find patterns, and give the business something it can act on.

This document is your investigation log.

---

## Investigation 1 — What Does a Typical Delivery Look Like?

Before asking what is wrong, you need to understand what normal looks like.

```{r overview-stats}
tibble(
  Metric = c("Total orders","Average delivery time","Median delivery time",
             "Fastest delivery","Slowest delivery","Standard deviation"),
  Value  = c(
    format(nrow(df), big.mark = ","),
    paste(round(mean(df$time_taken_min, na.rm = TRUE), 1), "min"),
    paste(round(median(df$time_taken_min, na.rm = TRUE), 1), "min"),
    paste(round(min(df$time_taken_min, na.rm = TRUE), 1), "min"),
    paste(round(max(df$time_taken_min, na.rm = TRUE), 1), "min"),
    paste(round(sd(df$time_taken_min, na.rm = TRUE), 1), "min")
  )
) |> kbl_nice("Delivery time at a glance")
```

```{r dist-plot, fig.cap="Delivery time distribution — most orders complete between 20 and 50 minutes"}
ggplot(df, aes(x = time_taken_min)) +
  geom_histogram(bins = 40, fill = "#01696f", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(df$time_taken_min, na.rm = TRUE),
             color = "#a12c7b", linetype = "dashed", linewidth = 1) +
  annotate("text",
    x = mean(df$time_taken_min, na.rm = TRUE) + 1.5, y = Inf,
    vjust = 2, hjust = 0,
    label = paste("Mean:", round(mean(df$time_taken_min, na.rm = TRUE), 1), "min"),
    color = "#a12c7b", size = 3.5) +
  coord_cartesian(xlim = c(10, 65)) +
  labs(title = "Distribution of Delivery Times",
       subtitle = "Where do most orders land?",
       x = "Delivery Time (minutes)", y = "Number of Orders") +
  theme_minimal(base_size = 13)
```

The distribution is roughly bell-shaped with a slight right skew. The right tail is where the customer complaints live.

> **Finding:** The problem is not the average. The problem is the tail. We need to find what pushes orders to the extreme end.

---

## Investigation 2 — Is It Just Distance?

```{r distance-scatter, fig.cap="Delivery time rises with distance — but the relationship is noisy"}
ggplot(df |> filter(!is.na(distance_km)), aes(x = distance_km, y = time_taken_min)) +
  geom_point(alpha = 0.12, color = "#01696f", size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, color = "#a12c7b", linewidth = 1.2) +
  coord_cartesian(ylim = c(5, 65), xlim = c(0, 25)) +
  labs(title = "Distance vs Delivery Time", subtitle = "Does farther always mean slower?",
       x = "Distance (km)", y = "Delivery Time (min)") +
  theme_minimal(base_size = 13)
```

```{r distance-bucket-table}
df |>
  mutate(dist_bucket = cut(distance_km,
    breaks = c(0, 5, 10, 15, 20, Inf),
    labels = c("0–5 km","5–10 km","10–15 km","15–20 km","20+ km"))) |>
  group_by(`Distance Range` = dist_bucket) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Std Dev` = round(sd(time_taken_min, na.rm = TRUE), 1),
            .groups = "drop") |>
  drop_na() |>
  kbl_nice("Average delivery time by distance — standard deviation stays high in every bucket")
```

> **Finding:** Distance explains the baseline, but not the variance. Something else is driving the extremes.

---

## Investigation 3 — Traffic Is the Real Multiplier

```{r traffic-boxplot, fig.cap="Traffic jam orders have a dramatically higher median and wider spread"}
df |>
  filter(!is.na(road_traffic_density)) |>
  mutate(road_traffic_density = factor(road_traffic_density,
    levels = c("Low","Medium","High","Jam"))) |>
  ggplot(aes(x = road_traffic_density, y = time_taken_min, fill = road_traffic_density)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.15) +
  scale_fill_manual(values = c(Low="#01696f",Medium="#d19900",High="#da7101",Jam="#a12c7b")) +
  labs(title = "Delivery Time by Traffic Condition",
       subtitle = "How much does congestion cost?",
       x = "Traffic Density", y = "Delivery Time (min)") +
  theme_minimal(base_size = 13) + theme(legend.position = "none")
```

```{r traffic-table}
low_t <- mean(df$time_taken_min[df$road_traffic_density == "Low"], na.rm = TRUE)
jam_t <- mean(df$time_taken_min[df$road_traffic_density == "Jam"], na.rm = TRUE)

df |>
  filter(!is.na(road_traffic_density)) |>
  mutate(road_traffic_density = factor(road_traffic_density,
    levels = c("Low","Medium","High","Jam"))) |>
  group_by(`Traffic Level` = road_traffic_density) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Median (min)`   = round(median(time_taken_min, na.rm = TRUE), 1),
            `% of All Orders`= paste0(round(100 * n() / nrow(df), 1), "%"),
            .groups = "drop") |>
  kbl_nice(paste("Traffic jams add ~",
    round(jam_t - low_t, 1), "extra minutes vs low-traffic orders"))
```

> **Finding:** Jam adds ~`r round(jam_t - low_t, 1)` minutes on average. For a platform promising 30-min delivery, this is the difference between a good review and a refund.

> **Business action:** Build dynamic ETAs that pull live traffic data. Warn customers proactively when a jam is on the rider's route.

---

## Investigation 4 — Pickup Delay: The Hidden Bottleneck

```{r pickup-dist, fig.cap="Most pickup delays are under 20 minutes — but outliers extend further"}
df |>
  filter(!is.na(pickup_delay_min), pickup_delay_min >= 0, pickup_delay_min <= 30) |>
  ggplot(aes(x = pickup_delay_min)) +
  geom_histogram(binwidth = 1, fill = "#a12c7b", color = "white", alpha = 0.85) +
  labs(
    title    = "Distribution of Pickup Delay",
    subtitle = "Most orders picked up within 10–15 minutes",
    x = "Pickup Delay (min)", y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r pickup-vs-delivery}
df |>
  filter(!is.na(pickup_delay_min), pickup_delay_min >= 0) |>
  mutate(delay_bucket = cut(pickup_delay_min,
    breaks = c(0, 5, 10, 15, 20, 30, Inf),
    labels = c("0–5","5–10","10–15","15–20","20–30","30+"))) |>
  group_by(`Pickup Delay (min)` = delay_bucket) |>
  summarise(Orders = n(),
            `Avg Total Delivery` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Median Total`       = round(median(time_taken_min, na.rm = TRUE), 1),
            .groups = "drop") |>
  drop_na() |>
  kbl_nice("Each pickup delay bucket directly raises total delivery time")
```

> **Finding:** Pickup delay propagates 1-to-1 into total delivery time. Unlike traffic, this is partly controllable.

> **Business action:** Track restaurants with consistently high prep delays. Dispatch riders only when food is nearly ready.

---

## Investigation 5 — When You Order Matters

```{r hourly-plot, fig.cap="Delivery times peak during morning and evening rush windows"}
df |>
  filter(!is.na(order_hour)) |>
  group_by(Hour = order_hour) |>
  summarise(Avg_Time = round(mean(time_taken_min, na.rm = TRUE), 1), .groups = "drop") |>
  ggplot(aes(x = Hour, y = Avg_Time)) +
  geom_line(color = "#01696f", linewidth = 1.2) +
  geom_point(color = "#a12c7b", size = 2.5) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Average Delivery Time by Hour of Day",
       subtitle = "When does the platform struggle most?",
       x = "Hour (24h format)", y = "Avg Delivery Time (min)") +
  theme_minimal(base_size = 13)
```

```{r rush-hour-table}
df |>
  filter(!is.na(is_rush_hour)) |>
  group_by(`Rush Hour?` = ifelse(is_rush_hour == 1, "Yes (7–9am, 12–2pm, 6–9pm)", "No")) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `% of Orders`    = paste0(round(100 * n() / nrow(df), 1), "%"),
            .groups = "drop") |>
  kbl_nice("Rush hour orders take longer on average")
```

> **Finding:** Dinner rush (6–9pm) combines peak volume with peak congestion — worst window.

> **Business action:** Pre-position riders near popular clusters 15–20 min before peak windows. Offer off-peak discounts to spread demand.

---

## Investigation 6 — Does City Type Change Everything?

```{r city-plot, fig.cap="Metro cities show the widest spread and highest median delivery times"}
df |>
  filter(!is.na(city)) |>
  ggplot(aes(x = city, y = time_taken_min, fill = city)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.15) +
  scale_fill_manual(values = c(
    "Metropolitian" = "#a12c7b", "Urban" = "#01696f", "Semi-Urban" = "#d19900")) +
  labs(title = "Delivery Time by City Type",
       subtitle = "Where is the platform underperforming?",
       x = NULL, y = "Delivery Time (min)") +
  theme_minimal(base_size = 13) + theme(legend.position = "none")
```

```{r city-table}
df |>
  filter(!is.na(city)) |>
  group_by(City = city) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Avg Distance`   = round(mean(distance_km, na.rm = TRUE), 1),
            `Avg Pickup Lag` = round(mean(pickup_delay_min, na.rm = TRUE), 1),
            .groups = "drop") |>
  arrange(desc(`Avg Time (min)`)) |>
  kbl_nice("Metro cities lag — high density creates congestion that outweighs infrastructure advantage")
```

> **Business action:** Deploy extra fleet in metro zones during evening hours. Consider geo-zone pricing to reflect true operational cost.

---

## Investigation 7 — Weather and Festival Days

```{r weather-plot, fig.cap="Fog and stormy conditions produce the highest and most variable delivery times"}
df |>
  filter(!is.na(weather_conditions)) |>
  ggplot(aes(x = reorder(weather_conditions, time_taken_min, median),
             y = time_taken_min, fill = weather_conditions)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.15) +
  coord_flip() +
  labs(title = "Delivery Time by Weather Condition",
       subtitle = "Adverse weather increases both average time and spread",
       x = NULL, y = "Delivery Time (min)") +
  theme_minimal(base_size = 13) + theme(legend.position = "none")
```

```{r festival-table}
df |>
  filter(!is.na(festival)) |>
  group_by(`Festival Day` = festival) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Avg Distance`   = round(mean(distance_km, na.rm = TRUE), 1),
            .groups = "drop") |>
  kbl_nice("Festival days drive order surges that stretch the fleet thin")
```

> **Business action:** Integrate weather APIs and public holiday calendars. Auto-trigger delay warnings on the app when adverse conditions are detected — do not let the app show a normal ETA during a thunderstorm.

---

## Investigation 8 — Does the Rider Make a Difference?

```{r ratings-scatter, fig.cap="Higher-rated riders tend to complete deliveries slightly faster"}
df |>
  filter(!is.na(delivery_person_ratings)) |>
  ggplot(aes(x = delivery_person_ratings, y = time_taken_min)) +
  geom_point(alpha = 0.12, color = "#006494", size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, color = "#a12c7b", linewidth = 1.2) +
  labs(title = "Rider Rating vs Delivery Time",
       subtitle = "Do experienced riders deliver faster?",
       x = "Rider Rating (out of 6)", y = "Delivery Time (min)") +
  theme_minimal(base_size = 13)
```

```{r vehicle-table}
df |>
  filter(!is.na(type_of_vehicle)) |>
  group_by(`Vehicle Type` = type_of_vehicle) |>
  summarise(Orders = n(),
            `Avg Time (min)` = round(mean(time_taken_min, na.rm = TRUE), 1),
            `Avg Distance`   = round(mean(distance_km, na.rm = TRUE), 1),
            .groups = "drop") |>
  arrange(`Avg Time (min)`) |>
  kbl_nice("Vehicle type influences speed especially in congested zones")
```

> **Business action:** Assign high-rated riders to long-distance orders. In metro jam zones, favour motorcycles over bicycles. Pull low-condition vehicles from active duty.

---

## What the Investigation Found

```{r summary-table}
tibble(
  `Delay Driver`  = c("Traffic (Jam)","Pickup delay","Rush hour","Metro congestion",
                      "Fog / Storm","Festival surges","Distance","Low rider rating"),
  `How It Hurts`  = c(
    paste(round(jam_t - low_t, 1), "extra min vs low traffic"),
    "Carries 1-to-1 into total time",
    "Demand + congestion compound",
    "Metro lags despite density",
    "Reduces speed, widens spread",
    "Demand surge strains fleet",
    "Direct linear effect",
    "Modestly slower on average"
  ),
  `Controllable?` = c(
    "Indirectly — routing + ETA comms",
    "Yes — dispatch timing + restaurant ops",
    "Yes — proactive rider positioning",
    "Partially — fleet reallocation",
    "No — communicate and prepare",
    "No — staff up in advance",
    "Partially — zone planning",
    "Yes — training + assignment"
  ),
  Priority = c("High","High","High","High","Medium","Medium","Medium","Low")
) |>
  kbl(caption = "What drives delays and what the business can do about it") |>
  kable_styling(full_width = TRUE, bootstrap_options = c("striped","hover")) |>
  column_spec(4, bold = TRUE,
    color = ifelse(c("High","High","High","High","Medium","Medium","Medium","Low") == "High",
      "#a12c7b", "#01696f"))
```

---

## What Comes Next

EDA has answered the first half of the brief — we know what drives delays, in which direction, and with what magnitude.

The second half is prediction. Can we estimate delivery time **before a rider is even dispatched**? That is what Phase 3 answers.

> [Continue to Phase 3 — Predictive Modelling →](03_modelling.qmd)

---

## Navigation

- [← Back to Preprocessing](01_preprocessing.qmd)
- [↑ Return to Home](../index.qmd)